---
title: Working with Census Data
author:
- ESR 512 - GIS in Geostatistics
- Postgraduate Institute of Science
- University of Peradeniya
- Thusitha Wagalawatta & Daniel Knitter
date: 10/2015
email: daniel.knitter@fu-berlin.de
bibliography: Course_Statistics_SriLanka.bib
csl: harvard1.csl
output:
  html_document:
    toc: false
    theme: united
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
  pdf_document:
    toc: false
    toc_depth: 2
    number_sections: true
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
highlight: pygments
---


## Census data ##
```{r setup, eval=TRUE, echo=FALSE}
setwd("/media/daniel/homebay/Teachings/WS_2015-2016/Statistics_SriLanka/")
```

First, we load the shapefile `Divsec` whose attributes contains the administrative areas on different levels for Sri Lanka.

```{r district_shp, eval=TRUE, echo=TRUE}
## load district shapfile
library(rgdal)
distr <- readOGR(dsn = "./data/DSL250-Shp/", layer = "Divsec")
plot(distr)
```
Since most of the data provided by LankaSIS is on the district level, we need to merge the present aggregation scheme on the district level. 

```{r distr_aggregate, eval=TRUE, echo=TRUE}
### solutions (see also: to http://gis.stackexchange.com/questions/63577/joining-polygons-in-r
dist.coords <- coordinates(distr)

# Generate IDs for grouping
distr.id <- distr@data$DISTRICT #cut(distr.coords[,1], quantile(distr.coords[,1]), include.lowest=TRUE)

# Merge polygons by ID
library(maptools)
distr.union <- unionSpatialPolygons(distr, distr.id)

## Plotting - option 1
par(mfrow = c(1,2))
plot(distr)
plot(distr.union)

## Plotting - option 2
## plot(distr)
## plot(distr.union, add = TRUE, border = "red", lwd = 2)

# Convert SpatialPolygons to data frame
distr.df <- as(distr, "data.frame")

## small function to find the mode
simpleMode <- function(x) {
	return(x[which.max(tabulate(x))])
	}

# Aggregate and sum desired data attributes by ID list
distr.df.agg <- aggregate(distr.df[, 5], list(distr.id), FUN = simpleMode)
row.names(distr.df.agg) <- as.character(distr.df.agg$Group.1)
head(distr.df.agg)

# Reconvert data frame to SpatialPolygons
distr.shp.agg <- SpatialPolygonsDataFrame(distr.union, distr.df.agg)

# Plotting
#library(gridExtra)
#grid.arrange(spplot(distr, "DISTRICT", main = "Distr: original county area"), 
#             spplot(distr.shp.agg, "Group.1", main = "Distr: aggregated county area"), ncol = 1)

str(distr.shp.agg@data)
names(distr.shp.agg@data) <- c("DISTRICT","PROVINCE")

writeOGR(distr.shp.agg, dsn = "./results/", layer = "districts", driver = "ESRI Shapefile", overwrite=TRUE)

```


```{r start-census, eval=TRUE, echo=TRUE}
## population by sector and district
## http://sis.statistics.gov.lk/statHtml/statHtml.do?orgId=144&tblId=DT_POP_SER_265&conn_path=I2
pop <- read.table(file = "./data/Pop.csv",
	header = TRUE,
	sep = ",", skip = 0, stringsAsFactors = FALSE)
str(pop)

## select entire population of districts and remove "Sri Lanka"
pop <- pop[!pop$Sri.lanka.standard.classification.of.area=="Sri Lanka" & pop$Sex=="Both Sexes" & pop$Sector=="All sectors",]
names(pop)[1] <- "DISTRICT"
pop <- subset(x = pop, select = c("DISTRICT", "X1971", "X1981", "X2001", "X2012"))
str(pop)
head(pop)

## fix the numbers
for (i in 2:5) {
    pop[,i] <- as.numeric(gsub(pattern = ",",replacement = "", x=pop[,i],fixed = TRUE))
}

## sort the census data
pop <- pop[order(pop$DISTRICT),]
##convert text so that it matches the shapefile
pop$DISTRICT <- toupper(pop$DISTRICT)

distr.shp.agg$DISTRICT %in% pop$DISTRICT
pop$DISTRICT[(distr.shp.agg$DISTRICT %in% pop$DISTRICT)==FALSE]
distr.shp.agg$DISTRICT[(distr.shp.agg$DISTRICT %in% pop$DISTRICT)==FALSE]
pop$DISTRICT[(distr.shp.agg$DISTRICT %in% pop$DISTRICT)==FALSE] <- as.character(distr.shp.agg$DISTRICT[(distr.shp.agg$DISTRICT %in% pop$DISTRICT)==FALSE])


library(dplyr)
distr.shp.agg@data <- left_join(distr.shp.agg@data, pop)

## plot choropleht map
## --------------------------------------------------
## https://cran.r-project.org/web/packages/tmap/vignettes/tmap-nutshell.html
## https://cran.r-project.org/web/packages/tmap/tmap.pdf
## install.packages("tmap", dependencies = TRUE)
library(tmap)

## quick plot
qtm(distr.shp.agg, fill="X2001")

## detailed plot
tm_shape(distr.shp.agg) +
    tm_fill("X2012", textNA="no data", title="Population 2012") +
    tm_text("DISTRICT", size= .5) +
    tm_borders() +
tm_layout(title = "Population Sri Lanka")


## Plot population development as Graph
## --------------------------------------------------
library(ggplot2)
a <- distr.shp.agg@data

library(reshape)
b <- melt(a, id.vars = c("PROVINCE", "DISTRICT"), measure.vars = c("X1971","X1981","X2001","X2012"))
str(b)
b$value <- b$value/1000
levels(b$variable) <- c("1971","1981","2001","2012")

ggplot(data = b, aes(x = variable, y = value, group = DISTRICT, col = DISTRICT)) +
    geom_line() +
    geom_point() +
    facet_grid(. ~ PROVINCE) +
    labs(title = "Population development Sri Lanka",
         x = "Census Year",
         y = "Population in thousands") +
    guides(col=guide_legend(nrow=5, keyheight=.5, keywidth=.5)) +
    theme(legend.position = "bottom",
	      text = element_text(size=10),
		  legend.text = element_text(size=8))


```

**Exercises**

1. Create maps for the population of the different provinces of Sri Lanka.

Tasks to perform:
- data aggregation
- Polygon union
- Plotting

2. Which province has the highest population standard deviation? Calculate and map your results.

3. Is there a significant difference between the sexes in the different districts? Calculate and map your results.



```{r census-sex, echo=TRUE, eval=TRUE}

pop <- read.table(file = "./data/Pop.csv",
	header = TRUE,
	sep = ",", skip = 0, stringsAsFactors = FALSE)
str(pop)
pop <- pop[!pop$Sri.lanka.standard.classification.of.area=="Sri Lanka" & !pop$Sex=="Both Sexes" & pop$Sector=="All sectors",]
names(pop)[1] <- "DISTRICT"
pop <- subset(x = pop, select = c("DISTRICT", "Sex","X1971", "X1981", "X2001", "X2012"))
str(pop)

## fix the numbers
for (i in 3:6) {
    pop[,i] <- as.numeric(gsub(pattern = ",",replacement = "", x=pop[,i],fixed = TRUE))
}

head(pop)


##  Poverty head count ratio (%) by Province and District (Poverty Head Count Ratio)
## http://sis.statistics.gov.lk/statHtml/statHtml.do?orgId=144&tblId=DT_LSS_POV_101&conn_path=I2
poverty <- read.table(file = "./data/poverty.csv", header = TRUE, sep = ",", skip = 0, stringsAsFactors = FALSE)
str(poverty)
str(pop)

## sex ratio
pop$X2012[pop$Sex=="Male"]/pop$X2012[pop$Sex=="Female"]

## is poverty and sex ratio related?
plot(pop$X2012[pop$Sex=="Male"]/pop$X2012[pop$Sex=="Female"], poverty$X2013)
cor(pop$X2012[pop$Sex=="Male"]/pop$X2012[pop$Sex=="Female"], poverty$X2013)

## is the distribution of sexes significantly different?



## does do ratio between the sexes change?
str(pop)
library(plyr)
a <- pop[pop$DISTRICT=="Colombo",]
a$X1971[2]

sr <- ddply(pop, "DISTRICT", function(x) {
                sr1 <- x$X1971[1]/x$X1971[2]
                sr2 <- x$X1981[1]/x$X1981[2]
                sr3 <- x$X2001[1]/x$X2001[2]
                sr4 <- x$X2012[1]/x$X2012[2]
                data.frame(sr.1971 = sr1,
                           sr.1981 = sr2,
                           sr.2001 = sr3,
                           sr.2012 = sr4)
            })

str(sr)

library(reshape)
sr2 <- melt(sr, id.vars = c("DISTRICT"), measure.vars = c("sr.1971","sr.1981","sr.2001","sr.2012"))
str(sr2)

ggplot(sr2, aes(x=variable, y=value, group=DISTRICT, col=DISTRICT)) +
    geom_line() +
        geom_point() +
            labs(title = "Sex-ratio development in Sri Lanka",
         x = "Census Year",
         y = "Sex ratio (Male/Female)") +
    guides(col=guide_legend(nrow=5, keyheight=.5, keywidth=.5)) +
    theme(legend.position = "bottom",
	      text = element_text(size=10),
		  legend.text = element_text(size=8))

## Join with district data
distr <- readOGR(dsn = "./results/", layer="districts")

## sort the census data
sr <- sr[order(sr$DISTRICT),]
##convert text so that it matches the shapefile
sr$DISTRICT <- toupper(sr$DISTRICT)
sr

distr$DISTRICT %in% sr$DISTRICT
sr$DISTRICT[(distr$DISTRICT %in% sr$DISTRICT)==FALSE]
distr$DISTRICT[(distr$DISTRICT %in% sr$DISTRICT)==FALSE]
sr$DISTRICT[(distr$DISTRICT %in% sr$DISTRICT)==FALSE] <- as.character(distr$DISTRICT[(distr$DISTRICT %in% sr$DISTRICT)==FALSE])

library(dplyr)
distr@data <- left_join(distr@data, sr)
str(distr@data)

spplot(obj = distr, zcol = c("sr.1971","sr.1981","sr.2001","sr.2012"))


## relation between sex ratio and mean elevation?
## H: the higher the elevation, the more tee production, the more women working
library(raster)
srtm <- raster("./results/srtm.tif")
distr@data$elev <- extract(x = srtm, y = distr, fun=mean, na.rm = TRUE)

cor(distr@data$elev,distr@data$sr.2012)



## Population by age group, sex, residential sector and district
## http://sis.statistics.gov.lk/statHtml/statHtml.do?orgId=144&tblId=DT_POP_GEO_174&conn_path=I2
## -> Pop 2001
pop2 <- read.table(file = "./data/Pop2.csv", header = TRUE, sep = ",", skip = 0, stringsAsFactors = FALSE)
str(pop2)
pop2 <- pop2[2:67,]
pop2$X2001 <- as.numeric(gsub(",","",pop2$X2001))
pop2$X2012 <- as.numeric(gsub(",","",pop2$X2012))
head(pop2)
str(pop2)

library(ggplot2)

ggplot(pop2, aes(x=District,y=X2001, fill=Residential.sector)) +
    geom_bar(stat = "identity") +
        coord_flip()


## Employed population by sex, province and district
## http://sis.statistics.gov.lk/statHtml/statHtml.do?orgId=144&tblId=DT_LFE_EMP_101&conn_path=I2
employed <- read.table(file = "./data/Employed_population.csv", header = TRUE, sep = ",", skip = 0, stringsAsFactors = FALSE)
str(employed)
employed$Male <- as.numeric(gsub(",","",employed$Male))
employed$Female <- as.numeric(gsub(",","",employed$Female))
names(employed)[2] <- "District"
head(employed)

library(reshape)
employed2 <- melt(employed, id.vars = c("District","Period"),measure.vars = c("Male","Female"))
str(employed2)
names(employed2)[3:4] <- c("Sex","Employed")

library(ggplot2)

ggplot(employed2, aes(x=District,y=Employed, fill = Sex)) +
    geom_bar(stat="identity") +
        coord_flip()


## Unemployment rate (%) by province and district
## http://sis.statistics.gov.lk/statHtml/statHtml.do?orgId=144&tblId=DT_LFE_UNP_101&conn_path=I2
unemployed <- read.table(file = "./data/Unemployment_rate.csv", header = TRUE, sep = ",", skip = 0, stringsAsFactors = FALSE)
str(unemployed)
tmp <- apply(X = unemployed[2:8], MARGIN = 2, FUN = function(x) {as.numeric(gsub(",","",x))})
head(tmp)
unemployed <- data.frame(DISTRICT = unemployed$Province.and.district, tmp)
str(unemployed)

library(reshape)
unemployed2 <- melt(unemployed, id.vars = "DISTRICT", measure.vars = names(unemployed[,2:8]))
str(unemployed2)
names(unemployed2)[2:3] <- c("Year","Unemployed")
head(unemployed2)

library(ggplot2)

ggplot(unemployed2, aes(x = Year, y = Unemployed, group = DISTRICT, col=DISTRICT)) +
    geom_line() +
        guides(col=guide_legend(nrow=5, keyheight=.5, keywidth=.5)) +
            theme(legend.position = "bottom",
          text = element_text(size=10),
          legend.text = element_text(size=8))


## Is there a significant difference in the employment of Male and Female?

## propotion of males employed
str(employed)
m.e.2012 <- employed2[employed2$Sex=="Male" & employed2$Period==2012,]
m.e.2012
str(pop)

library(reshape)
pop.r <- melt(pop, id.vars = c("DISTRICT", "Sex"), measure.vars = names(pop[3:6]))
str(pop.r)
names(pop.r) <- c("DISTRICT","Sex","Year","Population")
m.2012 <- pop.r[pop.r$Sex=="Male" & pop.r$Year=="X2012",]
m.2012

p.m.e.2012 <- data.frame(DISTRICT = m.e.2012[,1], pme = m.e.2012$Employed/m.2012$Population)
p.m.e.2012

## proportion of females employed
str(employed)
f.e.2012 <- employed2[employed2$Sex=="Female" & employed2$Period==2012,]
f.e.2012

f.2012 <- pop.r[pop.r$Sex=="Female" & pop.r$Year=="X2012",]
f.2012

p.f.e.2012 <- data.frame(DISTRICT = f.e.2012[,1], pfe = f.e.2012$Employed/f.2012$Population)
p.f.e.2012 

t.test(p.m.e.2012[,2],p.f.e.2012[,2]) #??


## Employed population by level of education, sex, province and district
## http://sis.statistics.gov.lk/statHtml/statHtml.do?orgId=144&tblId=DT_LFE_EMP_107&conn_path=I2
une.edu <- read.table(file = "./data/Employed_population_edu.csv", header = TRUE, sep = ",", skip = 0, stringsAsFactors = FALSE)
str(une.edu)
tmp <- apply(X = une.edu[,4:5], MARGIN = 2, FUN = function(x) {as.numeric(gsub(",","",x))})
head(tmp)
une.edu <- data.frame(YEAR = une.edu$Period,
                      DISTRICT = une.edu$Province.and.district,
                      LoE = une.edu$Level.of.education,
                      tmp)
str(une.edu)

library(reshape)
une.edu2 <- melt(une.edu, id.vars = c("YEAR","DISTRICT","LoE"), measure.vars = c("Male","Female"))
str(une.edu2)
names(une.edu2)[4:5] <- c("Sex","num_employed")
une.edu2[1,]

ggplot(une.edu2, aes(x=Sex, y=num_employed, fill=Sex)) +
    geom_bar(stat="identity") +
        facet_grid(LoE ~ DISTRICT)

## is employment rate linked to level of education?
a <- cbind(une.edu2[une.edu2$DISTRICT==levels(une.edu2$DISTRICT)[20] & une.edu2$Sex=="Male" & une.edu2$YEAR==2012,5],
           une.edu2[une.edu2$DISTRICT==levels(une.edu2$DISTRICT)[20] & une.edu2$Sex=="Female" & une.edu2$YEAR==2012,5]
           )
a
dimnames(a) <- list("LoE"=levels(une.edu2$LoE),
                    "Sex"=levels(une.edu2$Sex)
                    )
a
#barplot(a, beside = TRUE)
prop.table(a)
chisq.test(a) #??


```
