---
title: Point pattern
author:
- ESR 512 - GIS in Geostatistics
- Postgraduate Institute of Science
- University of Peradeniya
- Thusitha Wagalawatta & Daniel Knitter
date: 10/2015
email: daniel.knitter@fu-berlin.de
bibliography: Course_Statistics_SriLanka.bib
csl: harvard1.csl
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
highlight: pygments
---

# Point pattern analyses #

```{r setup, eval=TRUE, echo=FALSE}
setwd("/media/daniel/homebay/Teachings/WS_2015-2016/Statistics_SriLanka/")
library(raster)
```

_THE_ references for this subject are:

- [@bivand2008]
- [@baddeley2008]
- [@diggle2013]

And the best thing is: the authors are active developers of `R`. The package we will use in this topic very often is `spatstat`. A package that powerful that the manual alone has over 1400 pages (as of 2015-09-26).

If you use the package in your research, do not forget to cite it (this holds true for every package used, including the base software `R`):

```{r}
library(spatstat)
citation(package = "spatstat")
```

[following text largely from Knitter & Nakoinz (submitted)]

Point patterns are the result of processes that are influenced by (a) first-order effects, i.e. the location of the point is influenced by the underlying structure of the area but not by the location of other points [@wiegand2004, 210]; (b) second-order effects that occur when the location of a point is influenced by the presence or absence of other points [@wiegand2004, 210]. Point pattern analyses are common in ecological studies [e.g. @legendre2012; @wiegand2013]; up to now there are only few applications in archaeological contexts [e.g. @knitter2014].

## Warm up - some exploratory calculations ##

### Data preparation ###

Load shapefiles and create point dataset of the city polygons.

```{r data, eval=TRUE, echo=TRUE}
library(rgdal)
boundary <- readOGR(dsn = "./data/DSL250-Shp/", layer = "Boundary")
cities <- readOGR(dsn = "./data/DSL250-Shp", layer = "Cities")

library(rgeos)
cities.p <- gCentroid(cities, byid = TRUE)

par(mfrow = c(1,2))
plot(boundary)
plot(cities, add=TRUE)
plot(boundary)
plot(cities.p, add=TRUE)
```

### Mean center and standard distance ###

```{r mc-sd}

## prepare the data
x.co <- cities.p@coords[,1]
y.co <- cities.p@coords[,2]
n <- length(cities.p@coords[,1])

## calculate mean center
mc <- SpatialPoints(coords = cbind(sum(x.co)/n,sum(y.co)/n), proj4string = cities.p@proj4string)
mc

## calculate standard distance
sd <- sqrt(sum((x.co-mean(x.co))^2+(y.co-mean(y.co))^2)/n)
sd

## plot it
plot(boundary)
points(cities.p, pch = 19, cex = .5)
points(mc, col="red", pch = 19)
library(plotrix)
draw.circle(x = mc@coords[,1], y = mc@coords[,2], radius = sd, border = "red")
title("Mean center and standard distance \nof cities.p in Sri Lanka")
```

### Global intensity

```{r gintensity}
## get Area of Sri Lanka and change it to sqkm
area <- gArea(boundary)
area <- area/1000000

## calculate intensity
intensity <- length(cities.p)/area
intensity
```

> Exercise: what about the villages?!
>
> 1. Calculate mean centre and standard distance for the village dataset.
> 2. Plot the results
> 3. Calculate the global intensity of the village dataset. What does the value mean?

```{r ex1, eval=FALSE, echo=FALSE}
vil <- readOGR(dsn = "./data/DSL250-Shp/", layer = "Villname")
plot(vil)

mc2 <- SpatialPoints(coords = cbind(sum(vil@coords[,1])/length(vil),sum(vil@coords[,2])/length(vil)), proj4string = vil@proj4string)
mc2

sd2 <- sqrt(sum((vil@coords[,1]-mean(vil@coords[,1]))^2+(vil@coords[,2]-mean(vil@coords[,2]))^2)/length(vil))
sd2

plot(boundary)
points(vil, pch = 3, cex = .5)
points(mc, col="red", pch = 19, cex=2)
points(mc, col="blue", pch = 19)
library(plotrix)
draw.circle(x = mc@coords[,1], y = mc@coords[,2], radius = sd, border = "red")
draw.circle(x = mc2@coords[,1], y = mc2@coords[,2], radius = sd2, border = "blue")
title("Mean center and standard distance \nof cities.p (red) and villages (blue) in Sri Lanka")

area <- gArea(boundary)
area <- area/1000000

intensity2 <- length(vil)/area
intensity2
```

## Working with `ppp` objects -- Conducting Point Pattern Analysis ## 

First, let us create the point pattern. 

```{r create_pp}
## another window -- when "boundary" is too detailed
# library(mapdata)
# sl <- map("worldHires","Sri Lanka")
# sl <- data.frame(x=sl$x,y=sl$y)
# sl <- sl[is.na(sl$x)==FALSE & is.na(sl$y)==FALSE,]
# sl.owin <- owin(poly = sl)
# plot(sl.owin) ## quite rough, i know... but it's fast

library(spatstat)
library(maptools)
sl.owin <- as.owin.SpatialPolygons(unionSpatialPolygons(boundary, IDs = boundary@data$OUTERB_ID))
pp.cit <- ppp(x = cities.p@coords[,1], y = cities.p@coords[,2], window = sl.owin)
#str(pp.cit)
pp.vil <- ppp(x = vil@coords[,1], y = vil@coords[,2], window = sl.owin)

#plot(pp.vil, pch = 3, cex = .5)
plot(pp.cit, pch = 19, cols = "red")
```

As you can see the plot of `pp.cit` already shows the boundary of Sri Lanka. Why? And: why is this important? (hint: `?owin`)

### Quadrat count

```{r qc}
qc.cit <- quadratcount(X = pp.cit)
qc.vil <- quadratcount(X = pp.vil)

par(mfrow = c(1,2))
plot(qc.cit)
plot(qc.vil)
```

So the question is: does the quadratcount indicates CSR? To check we use a $\chi^2$ test approach (remember: Relation between observed (i.e. empirical) and expected (i.e. theoretical, here CSR) amounts of points in quadrants)

```{r qt}
qt.cit <- quadrat.test(X = pp.cit)
qt.cit
qt.vil <- quadrat.test(X = pp.vil)
qt.vil

# par(mfrow = c(1,2))
# plot(qt.cit)
# plot(qt.vil)
```

> **Exercise:** 
> 
> What influence does a change in the amount and density of the quadrants have? Is the current approach useful? 

### Nearest-neighbor distance

```{r nn}
nn.cit <- nndist(X = pp.cit)
nn.vil <- nndist(X = pp.vil)

str(nn.cit)

mean(nn.cit)
mean(nn.vil)

hist(nn.cit)
abline(v=mean(nn.cit))
abline(v=median(nn.cit), lty=2)

hist(nn.vil)
abline(v=mean(nn.vil))
abline(v=median(nn.vil), lty=2)
```

### Nearest neighbor Distance -- Clark and Evans’ R

"An R value of less than 1 indicates of a tendency toward clustering, since it shows that observed nearest neighbor distances are shorter than expected. An R value of more than 1 indicatives of a tendency toward evenly spaced events" (O'Sullivan & Unwin 2010, 144) 

```{r clarkevans}
nnE <- 1/(2*sqrt((pp.vil$n/gArea(boundary))))
nnE
R.vil <- mean(nn.vil)/nnE
R.vil

nnE <- 1/(2*sqrt((pp.cit$n/gArea(boundary))))
nnE
R.cit <- mean(nn.cit)/nnE
R.cit
```

## First order effects ##

### Density calculation - 1st approach ###

Amount of events within pixel...in imprecise language: a histogram in space. This is a step by step example but the code is rather inefficient. 

```{r density1}
cs <- 50000    # cellsize
# enlarge the study area by ONE pixel (in E-W and N-S direction)
xmin <- pp.cit$window$xrange[1] - cs/2 # enlarge the study area by cs/2
xmax <- pp.cit$window$xrange[2] + cs/2 # enlarge the study area by cs/2
ymin <- pp.cit$window$yrange[1] - cs/2 # enlarge the study area by cs/2
ymax <- pp.cit$window$yrange[2] + cs/2 # enlarge the study area by cs/2
rows  <- round((ymax-ymin)/cs, 0) + 1 # calculate the number of rows (add 1 just because a pixl might get lost through the rounding operation)
columns <- round((xmax-xmin)/cs, 0) + 1 # calculate the number of columns (add 1 just because a pixl might get lost through the rounding operation)
z <- cbind(1:(columns*rows)) # create a vector with all grid cells
df <- data.frame(z) # create a data.frame of it
gt <- GridTopology(c(pp.cit$window$xrange[1] - cs/2,pp.cit$window$yrange[1] - 
                     cs/2), c(cs,cs), c(columns,rows)) # create a topological description of a grid and aftecsards...
gt # ...have a look at it and...
sgdf <- SpatialGridDataFrame(gt, df, proj4string = cities.p@proj4string) # ...create the grid.

for (i in seq(along=coordinates(gt)[,1])){ # loop for every cell
    x <- coordinates(gt)[i,1] - cs/2 # because the coordinate is define for the center of the cell, the half cellsize is substracted
    y <- coordinates(gt)[i,2] - cs/2 # because the coordinate is define for the center of the cell, the half cellsize is substracted
    xi <- which(pp.cit$x>x & pp.cit$x<x+cs) # which events lie within the x direction?
    yi <- which(pp.cit$y>y & pp.cit$y<y+cs) # which events lie within the y direction? 
    pz <- length(intersect(xi,yi)) # how many objects in x and y direction intersect?
    sgdf@data$z[i]<- pz / (cs/1000)^2  # divide the number of points by the area
}

plot(raster(sgdf), col = gray.colors(25, start = 0.97, end = 0.4))      
plot(boundary, add=TRUE)
points(pp.cit$x, pp.cit$y, pch=16, cex=0.4)  
```


### Density calculation - 2nd approach ###

Kernel density estimation; again here performed step by step and rather inefficiently coded. Use it to understand but do not run the code. 

```{r density2}
sgdf_kde <- sgdf # copy the grid into a new variable
sd <- 100000 # define the bandwidth of the kernel - it is NOT the pixel size; this is defined in the beginning of the 1str approach
# ...and think of it, the bandwidth shall not be smaller than the pixelsize, cause points outside the kernel are "underestimated"

for (i in seq(along=coordinates(gt)[,1])){ # loop through all pixel
    x <- coordinates(gt)[i,1]
    y <- coordinates(gt)[i,2]
    g2 <- 0
    for (j in seq(along=pp.cit$x)){ # we use ALL points
        distanz <- sqrt((pp.cit$x[j] - x)^2 + (pp.cit$y[j] - y)^2) # calculate this distance of one point to the centre of the raster cell. this is necessary for the...
        g1 <- dnorm(distanz, mean=0, sd=sd) #...weighting of point in th e cell using a normal distributed kernel
        g2 <- g2 + g1 # now collect the weights for the different points under in the pixel to get the density value for the pixel
    }
    sgdf_kde@data$z[i]<- g2     
}
plot(raster(sgdf_kde), col = gray.colors(25, start = 0.97, end = 0.4))   
plot(boundary, add=TRUE)
points(pp.cit$x, pp.cit$y, pch=16, cex=0.4) 
```

### Density calculation - 3rd approach ###

Kernel density estimation using the built in function of `spatstat`. Since the package is great, it is perfectly coded and fast. Play around with the values. Calculate also a KDE for the village data set. 

```{r density3}
cs <- 1000  
sd <- 50000

dens_r5 <- density(pp.cit, sd, eps=cs, edge=TRUE, at="pixels") 
plot(dens_r5, col = gray.colors(25, start = 0.97, end = 0.4))
#contour(dens_r5, add=T)    
points(pp.cit$x, pp.cit$y, pch=16, cex=0.4)  

# use another measure for sd - in this case three times the mean nearest neighbor distance
sdev <- 3*mean(nndist(pp.cit))  
dens_r6 <- density(pp.cit, sdev, eps=cs, edge=TRUE, at="pixels")
plot(dens_r6, col = gray.colors(25, start = 0.97, end = 0.4))
#contour(dens_r6, add=T)    
points(pp.cit$x, pp.cit$y, pch=16, cex=0.4)  
```

> **Exercise**
>
> 1. Use the `density` function to calculate the KDE for the village data set. 
> 2. Change the size of the kernel and interpret/discuss the results
> 3. Get familiar with the function and concepts using the `spatstat` manual

## Second order effects ##

### G and F function

First approach based on theoretical assumptions, i.e. CSR

```{r GF1}
g.cit <- Gest(pp.cit)
plot(g.cit)

f.cit <- Fest(pp.cit)
plot(f.cit)
```

Second approach based on simulations of a theortical process, i.e. CSR

```{r GF2}
g.cit.env <- envelope(pp.cit, fun = Gest, nsim = 10)
plot(g.cit.env)

f.cit.env <- envelope(pp.cit, fun = Fest, nsim = 10)
plot(f.cit.env)
```

> **Question:** Interpret the results. Why two approaches? What are the advantages?

### K and L function

First approach based on theoretical assumptions, i.e. CSR

```{r kl1}
k.cit <- Kest(pp.cit)
plot(k.cit)

l.cit <- Lest(pp.cit)
plot(l.cit)
```

Second approach based on simulations of a theortical process, i.e. CSR

```{r kl2}
k.cit.env <- envelope(pp.cit, fun = Kest, nsim = 10)
plot(k.cit.env)

l.cit.env <- envelope(pp.cit, fun = Lest, nsim = 10)
plot(l.cit.env)
```

> **Question:** Interpret the results. Why two approaches? What are the advantages?

> **Exercise**
>
> 1. Calculate G,F,K, and L function for the village data set
> 2. Interpret the results. Which approach do you chose? Why?
> 3. Get familiar with the function and concepts using the `spatstat` manual

# References #

