---
title: Viewshed Analyses
author:
- ESR 512 - GIS in Geostatistics
- Postgraduate Institute of Science
- University of Peradeniya
- Thusitha Wagalawatta & Daniel Knitter
date: 10/2015
email: daniel.knitter@fu-berlin.de
bibliography: Course_Statistics_SriLanka.bib
csl: harvard1.csl
output:
  html_document:
    toc: false
    theme: united
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
  pdf_document:
    toc: false
    toc_depth: 2
    number_sections: true
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
highlight: pygments
---

```{r setup, eval=TRUE, echo=FALSE}
setwd("/media/daniel/homebay/Teachings/WS_2015-2016/Statistics_SriLanka/")
```

## Viewshed Analysis ##

Be aware, for the following to work you need to have GRASS GIS installed on your system.

Load library, initialize the GRASS environment with default values (everything is stored as temporary files; you can change this if you want to come back to earlier stages of your work)

```{r init_grass, eval = TRUE, echo = TRUE}
library(spgrass6)
loc <-  initGRASS("/usr/local/grass-7.0.0svn/", home=tempdir() ,mapset = "PERMANENT", override = TRUE)
execGRASS("g.proj", flags = c("c"), parameters = list(proj4="+init=epsg:32644"))

library(rgdal)
srtm <- readGDAL("./results/srtm.tif")
writeRAST6(x = srtm, vname="dem", flags = c("overwrite"))
## adjust Regions resolution!
execGRASS("g.region", parameters = list(raster = "dem",res = as.character(srtm@grid@cellsize[1])))
execGRASS("g.region", flags = c("p"))
```

Now we are ready to perform our viewshed analysis. 

```{r eval = TRUE, echo = TRUE}
# display the possible commands in grass
parseGRASS("r.viewshed")
```

Let us create viewshed maps for the peak data

```{r eval = TRUE}
peaks <- readOGR("./data/DSL250-Shp", "Peaks")
library(raster)
peaks <- spTransform(peaks, raster(srtm)@crs)

## for better orientation insert the boundary
boundary <- readOGR("./data/DSL250-Shp", "Boundary")
boundary <- spTransform(boundary, raster(srtm)@crs)
```

```{r eval = TRUE, results="hide", warning = FALSE}
plot(boundary)
points(peaks, pch = 19, cex = .5)

## crop srtm to be relevant for dagoba data
# srtm_crop <- crop(raster(srtm), rbind(c(min(dagoba@bbox[1,]-10000),
#                                 max(dagoba@bbox[1,]+10000)),
#                               c(min(dagoba@bbox[2,]-10000),
#                                 max(dagoba@bbox[2,]+10000))
#                               ))		

srtm_crop <- raster(srtm)
writeRAST6(x = as(srtm_crop,"SpatialGridDataFrame"), vname="dem_crop", flags = c("overwrite"))
execGRASS("g.region", parameters = list(raster = "dem_crop"))

## viewshed for one point
co.peaks <- peaks@coords
execGRASS("r.viewshed", flags = c("overwrite","b"), parameters = list(input = "dem_crop",output = "view.peak",coordinates = co.peaks[1,]))
single.viewshed <- readRAST6("view.peak")
plot(raster(single.viewshed))
points(peaks[1,])
```

> **Question** 
> What might be wrong with this result?

```{r eval = TRUE, results="hide", warning = FALSE}
execGRASS("r.viewshed", flags = c("overwrite","b"), parameters = list(input = "dem_crop",output = "view.peak",coordinates = co.peaks[1,], max_distance=50000))
single.viewshed <- readRAST6("view.peak")
plot(raster(single.viewshed))
points(peaks[1,],pch=19)
```


```{r view_loop, eval=TRUE, results="hide", warning = FALSE}
## viewshed for all points
## ------------------------------
## load basic raster from grass
dem_crop <- readRAST6(vname = "dem_crop")

## loop through all points, calculate viewshed, and write in a an raster brick object
cum.view.peaks <- brick(raster(dem_crop))

for (i in seq(1, length(co.peaks[,1]))) {
    execGRASS("r.viewshed"
              ,flags = c("overwrite","b")
              ,parameters = list(input = "dem_crop",
                                 output = "view.peak",
                                 coordinates = co.peaks[i,],
                                 max_distance=75000)
              )
    viewshed <- readRAST6("view.peak")
    cum.view.peaks[[i]] <- raster(viewshed)
  	names(cum.view.peaks[[i]]) <- as.character(peaks$NAME[i])
    cat("iteration ", i, " of ", length(co.peaks[,1]),"\n")
}

plot(cum.view.peaks)

view.sum.peaks <- sum(cum.view.peaks)
plot(view.sum.peaks)

quantile(view.sum.peaks)
##   0%  25%  50%  75% 100% 
##   0    0    1   2  9 
##   1    2    3   4   5
rcl.peaks <- c(-Inf, quantile(view.sum.peaks)[2],1,
         quantile(view.sum.peaks)[2],quantile(view.sum.peaks)[3],2,
         quantile(view.sum.peaks)[3],quantile(view.sum.peaks)[4],3,
         quantile(view.sum.peaks)[4],quantile(view.sum.peaks)[5],4,
         quantile(view.sum.peaks)[5],+Inf,5
         )
rcl.peaks <- matrix(rcl.peaks, ncol = 3, byrow = TRUE)
view.sum.peaks2 <- reclassify(x = view.sum.peaks, rcl = rcl.peaks)
plot(view.sum.peaks2)
points(peaks, pch=19, cex=.5)

#library(rasterVis)
#levelplot(view.sum.peaks2)
```
